---
title: "PacSci Learn at home (Under Construction)"
output: 
  learnr::tutorial:
    css: "css/styles.css"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(conflicted)
library(tidyverse)
library(plotly)
library(ggsci)

conflict_prefer("filter", "dplyr")

knitr::opts_chunk$set(echo = FALSE)
d <- read_rds(path = "./pacSci__TCGA_BRCA_data.RDS")

```

## Introduction

### A Short Video
![](https://www.youtube.com/watch?v=Ct18nkmIFaw)

### Summary Stats
As with any analysis, we first need to better understand our data. Some basic descriptive statistics can help us evaluate how representative our sample is to the population for which we are making an inference.

```{r data overview, fig.width=10}
d$clin %>% group_by(race) %>% tally %>% ggplot(aes(x="",y=n,fill=race)) + 
  geom_col(width = 1) +  coord_polar("y", start=0) + theme_void() + 
  ggsci::scale_fill_jama() + theme(legend.position = "bottom")

(d$clin %>% filter(!is.na(age_at_diagnosis)) %>% ggplot(aes(x=age_at_diagnosis/365)) + 
    geom_histogram(colour="white",bins = 30) + xlab("Age at diagnosis (years)")) %>%
  ggplotly(.)
```




## Topic 1

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

